% Introduction section — DESI paper (Round 2 revision)
% Fixed: PSMC/MSMC differentiation; softened "falsifying"; removed Tajima1983 misuse;
%        Para 5 restructured (insight first, methods secondary); Para 6 shortened/softened

\section*{Introduction}

Whether the ancestral population of modern humans was panmictic or structured during the Middle Pleistocene has profound implications for understanding human origins. The Early to Middle Pleistocene transition (MPT; $\sim$1.25–0.70 Ma) remains deeply contested: it has almost no fossil record on most continents, yet it encompasses the last common ancestor of modern humans, Neanderthals, and Denisovans \citep{Stringer2016}. The emergence of anatomically modern \textit{Homo sapiens} in Africa by $\sim$300 ka and its subsequent global dispersal represent a well-studied chapter \citep{Hublin2017}; the million years that precede it do not. Genomic data offer a principled window into this earlier period, because the distribution of pairwise coalescence times across the genome encodes ancestral population size, structure, and connectivity over millions of years — if the right statistical framework is applied.

The past decade has produced a series of increasingly powerful methods for inferring effective population size ($\Ne$) over time. Pairwise sequential Markovian coalescent approaches infer $\Ne(t)$ from pairs of individual genome sequences \citep{Li2011,Schiffels2014}. SFS-based and coalescent-aware methods scale this to larger samples or exploit allele-frequency distributions \citep{Gutenkunst2009,Terhorst2017}, and composite likelihood approaches extend the framework to extremely large cohorts \citep{Hu2023}. Despite their diversity, these approaches share a critical unexamined assumption: that the ancestral population was panmictic — a single, randomly mating entity with no internal structure. Even the pairwise coalescent methods (PSMC, MSMC2) estimate a single $\Ne(t)$ trajectory per individual and do not provide a statistical test of panmixia versus structure. Although PSMC-based analyses do show greater diversity in African individuals than in non-Africans \citep{Malaspinas2016,Mallick2016}, this per-individual depth difference is expected under both panmictic and structured models, and does not test whether the \textit{difference in within-group coalescence time} across continental populations is compatible with panmixia. That explicit test had not previously been performed.

This assumption was thrown into sharp relief by \citet{Hu2023}, who used FitCoal to infer a catastrophic population bottleneck at $\sim$930 ka, reducing ancestral human $\Ne$ to just $\sim$1,280 individuals for approximately 117,000 years — a finding interpreted as a genomic signature of the MPT. Two subsequent studies challenged the interpretation. \citet{Cousins2025b} reanalysed the SFS and found that a panmictic model \textit{without} a severe bottleneck fits the data substantially better. More fundamentally, \citet{Cousins2025} applied Cobraa — a structured coalescent hidden Markov model — and inferred two ancestral populations diverging at $\sim$1.5 Ma with $\sim$20\% gene flow at $\sim$300 ka, providing a structured alternative that does not require a panmictic bottleneck. These three studies now present the field with three qualitatively different pictures of the same evolutionary interval, with no consensus.

The root difficulty is methodological: SFS-based inference under a panmictic model conflates ancestral population structure with fluctuations in effective population size. A structured ancestral population, in which genealogical lineages are partitioned across distinct demes, generates the same spectrum of rare and common variants as a transiently smaller panmictic population \citep{Cousins2025b}. Cobraa addresses this by fitting a parametric structured model, but does so via a complex hidden Markov model that requires full-genome phased sequences and is sensitive to the assumed model topology. What has been lacking is a direct, model-light statistical test of the panmixia null itself — one that does not commit to a specific structured demography.

Here we introduce \desi{} (Deep-time Evolutionary Structure Inference), which exploits a simple but previously untested prediction. Under panmixia, all modern continental groups are random draws from the same ancestral genealogy; consequently, within-group mean pairwise TMRCA must be equal across groups up to sampling variance — regardless of bottleneck parameterization. Ancestral population structure, by contrast, produces persistent genealogical depth differences between groups, because populations inheriting from distinct ancestral demes carry different coalescent histories. \desi{} tests this prediction directly by computing group-level mean pairwise TMRCA from per-window aggregate heterozygosity counts (100 kb windows, Poisson likelihood), with uncertainty from block jackknife over 1 Mb genomic blocks. This approach requires no assumptions about demographic model topology, provides a quantitative null (panmixia predicts zero difference) alongside a Cobraa-inspired reduced-$\Ne$ parameterization that predicts a specific difference matched by simulation (370.1 ka), and is validated to $<3.5\%$ calibration error on msprime coalescent simulations across three distinct demographic models.

Applying \desi{} to 240 individuals from the 1000 Genomes Project \citep{Byrska2022}, we find that within-group mean TMRCA differs substantially and consistently across continental populations: within-AFR ($1{,}229.5 \pm 6.4$ ka) exceeds within-EAS ($858.8 \pm 5.4$ ka) by $370.7 \pm 8.4$ ka ($Z = 73$ across all 22 autosomes). Coalescent simulations reveal a three-way comparison: a strictly panmictic model (single undivided population) predicts $\sim$0.2 ka; a standard Out-of-Africa model (deep panmixia + 60 ka divergence) predicts 463.9 ka — \emph{exceeding} the observation; and a Cobraa-inspired model with reduced ancestral $\Ne$ predicts 370.1 ka, matching within 0.2\%. The per-window TMRCA distribution shows no substantial fraction of East Asian windows coalescing below 100 ka, inconsistent with a simple Out-of-Africa bottleneck model that would require such a fraction to explain the observed mean. The signal is preserved with equal magnitude in genomic windows more than 50 kb from annotated genes, arguing against background selection as a confound. Together, these results provide independent genealogical evidence consistent with ancestral population structure and inconsistent with the panmictic bottleneck interpretation of the $\sim$930 ka signal.
